INLA replication

### 2022-10-24 src/example_inla-replication/inla-replication.Rmd
Improve this a bit

---
title: "INLA replication"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    | We demonstrate a step-by-step example of the INLA method implemented in R and TMB.
---

@@

Consider a three-stage model
\begin{alignat}{2}
  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \, 0,  Q(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
\end{alignat}
with posterior
\begin{equation}
  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
\end{equation}

@@

Using `TMB` if we integrate out the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:

@@

$K = 1$ with integration point $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
This can be found by minimising `h` using optimisers like `nlminb` or `optim`.

### INLA grid

The steps required are:

1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
2. Calculate negative Hessian at the mode using e.g. `h$env$spHess`.
3. Reparameterise in terms of $z$ coordinates along the eigenvectors of the Hessian starting at the mode
3. Test points at a certain step-size ($\delta_z$) against a condition whereby they must have sufficient posterior density (using a parameter $\delta_{\pi}$)

For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/elgm-inf/inla-grid.html) notebook.

### Adaptive Gaussian Hermite quadrature

We could also use an adaptive Gauss Hermite quadrature rule.
See the [Implementing Approximate Bayesian Inference using Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.

@@

Obtain marginals of $\tilde p_{\text{G}}(x \, | \, \theta, y)$ from above.

@@

One of the INLA vignettes discusses how the user can set their own integration points -- see `browseVignettes(package = "INLA")`.
The integration points that `R-INLA` uses can be found using `fit$joint.hyper`, which we can add on to the above plot for interest:

@@

# Different normalising constant?
ggplot(dense_theta) +
  geom_line(aes(x = input, y = output), col = midblue) +
  geom_point(data = internals_inla_theta, aes(x = input, y = output), shape = 4) +
  theme(axis.text.y = element_blank()) +
  labs(title = "INLA integration points for the hyperparameters") +
  theme_minimal()
```

@@

```{r}
blangiardo_x <- seq(-8, 5, length.out = 50)
```

@@

Following the book:

```{r}
nl_post_marginal_x <- rowSums(nl_joint_post)
plot(blangiardo_x, nl_post_marginal_x)
blangiardo_x[[which.min(nl_post_marginal_x)]]
```

@@

Compare $x$:

>>>>

---
title: "INLA replication"
author:
- name: Adam Howes
output:
  html_document:
    toc: yes
    toc_float: true
    toc_collapsed: true
    df_print: paged
    code_folding: show
    theme: lumen
abstract: |
    | We demonstrate a step-by-step example of the INLA method implemented in R and `TMB`.
---

@@

Consider a three-stage model
\begin{alignat}{2}
  &\text{(Observations)}     &   y_i &\sim p(y_i \, | \, x_i, \theta), \quad i \in \mathcal{I}, \\
  &\text{(Latent field)}     &   x &\sim \mathcal{N}(x \, | \, 0,  Q(\theta)^{-1}), \\
  &\text{(Hyperparameters)}  &   \qquad \theta &\sim p(\theta), \label{eq:hyper}
\end{alignat}
with posterior distribution
\begin{equation}
  p(x, \theta \, | \, y) \propto p(y, x, \theta) = p(y \, | \, x, \theta) p(x \, | \, \theta) p(\theta).
\end{equation}

@@

Using `TMB` if we integrate out a Gaussian approximation to the latent field $x$ using the argument `random = "x"` in `MakeADFun` then the result is the negative log of the Laplace approximation to the marginal likelihood $h(\theta) \equiv - \log L^\star(\theta)$:

@@

$K = 1$ with integration point $\hat \theta = \text{argmin}_{\theta} (-\log L^\star_f(\theta))$.
This can be found by minimising `h` using optimisers like `nlminb` or `optim`.
Sometimes methods like this are referred to as "[Empirical Bayes](https://en.wikipedia.org/wiki/Empirical_Bayes_method)" where the hyperparameters are estimated with a fixed point estimate.

### `R-INLA` grid method

Instead of just taking a single point, we might want to explore the posterior approximation to $p(\theta \, | \, y)$ more fully.
In `R-INLA` a few choices are available as to how to do this, including the grid integration strategy as follows:

1. Locate the mode of $\log p(\theta \, | \, y)$ which is the same as $\text{argmin}_{\theta} (-\log L^\star_f(\theta))$ and can be calculated as above.
2. Calculate negative Hessian at the mode (this could be done in `TMB` using h$env$spHess`)
3. Reparameterise in terms of $z$ coordinates along the eigenvectors of the Hessian starting at the mode
4. Test points at a certain step-size ($\delta_z$) against a condition whereby they must have sufficient posterior density (using a parameter $\delta_{\pi}$)

Under this approach, all weights would be equal $\Delta_k = 1 / K$.
For a more detailed run through of the grid integration method, see the [INLA two dimensional grid replication](https://athowes.github.io/elgm-inf/inla-grid.html) notebook.

### Adaptive Gauss-Hermite quadrature

We could also use an adaptive Gauss-Hermite quadrature (AGHQ) rule.
See the [Implementing Approximate Bayesian Inference using Adaptive Quadrature: the `aghq` Package](https://arxiv.org/pdf/2101.04468.pdf) paper.
AGHQ is attractive because it is automatic, works well for integrating densities close to being Gaussian.

## Stage 3: get $\tilde p(x_i \, | \, \theta, y)$

@@

The simplest approach would be to directly obtain the marginals of $\tilde p_{\text{G}}(x \, | \, \theta, y)$ from above.

@@

One of the INLA vignettes discusses how the user can set their own integration points -- see `browseVignettes(package = "INLA")`.
The integration points that `R-INLA` uses can be found using `fit$joint.hyper`:

@@

internals_inla_theta$input
```

@@

```{r}
x_start <- -8
x_end <- 5
x_out <- 50
blangiardo_x <- seq(x_start, x_end, length.out = x_out)
```

@@

Calculating the log posterior by integrating over the hyperparameters for each value of $x$:

```{r}
K <- length(points)
nl_post_marginal_x <- rowSums(nl_joint_post) / K

df <- data.frame(
  x = blangiardo_x,
  log_post = -1 * nl_post_marginal_x
)

ggplot(df, aes(x = x, y = log_post)) +
  geom_line() +
  theme_minimal() +
  labs(x = "x", y = "log(Posterior)")
```

Try normalising:

```{r}
x_spacing <- (x_end - x_start) / x_out
safe_exp <- function(x) exp(x - max(x))
df$post <- safe_exp(df$log_post)
norm <- trapezoid_rule(df$post, x_spacing)
df$norm_post <- df$post / norm

ggplot(df, aes(x = x, y = norm_post)) +
  geom_line() +
  theme_minimal() +
  labs(x = "x", y = "Posterior")
```

@@

What about for $x$:
